---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`optithresholdr` is designed for studies that use repeated wearable
measurements and want interpretable time-in-range summaries. A typical example
is CGM data, where each participant contributes many glucose readings and the
goal is to choose thresholds in a mathematically grounded way rather than
relying only on conventional fixed cutoffs.

The central idea is to choose thresholds so that the resulting time-in-range
summary optimally preserves distributional information from the original
measurement profiles.

The package has three user-facing functions:

- `as_distribution()` preprocesses repeated subject-level measurements
- `evaluate_loss()` scores a candidate cutoff set
- `optimal_thresholds()` estimates cutoffs with specified `loss1` or `loss2`

The recommended workflow is simple:

1. Start from a list of subject vectors or a long data frame.
2. Convert the data with `as_distribution()`.
3. Use `evaluate_loss()` if you want to check a specific cutoff set.
4. Use `optimal_thresholds()` if you want the package to estimate the cutoffs.

The package uses the full measurement profiles across participants when fitting
thresholds, aiming to maximally preserve distributional information.

`loss1` and `loss2` answer different questions. Use `loss1` when you want 
the resulting time-in-range summaries to stay close to each participant's
original measurement profile. Use `loss2` when you want the thresholded
summaries to preserve between-participant differences, which is often useful in
mixed or heterogeneous cohorts.

Missing values are removed before subject distributions are formed, and
subjects with no observed values left are dropped. For data frames, `id` and
`value` are supplied as strings.

## Data Frame Input

If your data are stored in long format, pass the participant and measurement
columns as strings.

```{r}
df <- data.frame(
  person = rep(c("A", "B", "C"), c(5, 6, 6)),
  reading = c(90, 110, 130, 150, 180,
              75, 82, 95, 105, 125, 160,
              60, 70, 88, 92, 115, 140)
)

distribution_df <- as_distribution(
  df,
  id = "person",
  value = "reading",
  range = c(40, 400),
  M = 200
)
```

Once the distribution object is created, the downstream workflow is the same as
for list input.

## A Small Example

```{r}
library(optithresholdr)

subjects <- list(
  subject_1 = c(90, 110, 130, 150, 180),
  subject_2 = c(75, 82, 95, 105, 125, 160),
  subject_3 = c(60, 70, 88, 92, 115, 140)
)

distribution <- as_distribution(subjects, range = c(40, 400), M = 200)

distribution
```

`as_distribution()` creates the internal object used by the loss and
optimization functions. In practice, this is the preprocessing step you run
once before scoring or estimating thresholds.

## Estimating Thresholds

To estimate thresholds, choose the number of free thresholds `K` and the loss. The
returned cutoffs optimize the associated time-in-range summaries.

```{r}
fit_loss1 <- optimal_thresholds(
  distribution,
  K = 2,
  loss = "loss1",
  seed = 123
)

fit_loss2 <- optimal_thresholds(
  distribution,
  K = 2,
  loss = "loss2",
  seed = 123,
  control = list(NP = 30, itermax = 100, trace = FALSE)   # DEOptim parameters can be specified
)

fit_loss1
fit_loss2
```

An `optithreshold_fit` object contains:

- `cutoffs`: the fitted interior thresholds
- `objective`: the achieved loss value
- `loss`: the loss definition used for fitting
- `wdist`: the base Wasserstein distance used inside the loss
- `elapsed`: elapsed optimization time in seconds

The printed summary is intentionally short, so the main quantities to inspect
are usually `fit$cutoffs` and `fit$objective`.

## Fixed Thresholds (semisupervised)

Sometimes part of the threshold set is fixed by study convention and you only
want to estimate the remaining cutoffs. Use the `fixed` argument for that case.

```{r}
fit <- optimal_thresholds(
  distribution,
  K = 1,        # additional thresholds to optimize
  fixed = 70,
  loss = "loss1",
  seed = 1,
  control = list(NP = 20, itermax = 50, trace = FALSE)
)

fit
```

Here `K = 1` means that one additional cutoff is optimized while `70` is kept
fixed in the final set.

## Real CGM data example with iglu




## Solver Notes

The optimizer uses `DEoptim` together with some parts of loss evaluation
accelerated with Rcpp.

The `control` argument is passed through to `DEoptim::DEoptim.control()`. The
package defaults are `NP = max(4, 15 * K)`, `itermax = 1000`, `CR = 0.7`,
`F = 0.75`, `strategy = 3`, `reltol = 1e-4`, `steptol = 30`, `bs = FALSE`,
`trace = FALSE`, and `parallelType = "none"`. If you supply
`control$initialpop`, its row count becomes the effective `NP`.
