---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(OptiThresholdR)
```

## Overview

`OptiThresholdR` finds data-driven thresholds that partition wearable device
measurements into time-in-range (TIR) proportions. The thresholds are chosen so
that the resulting TIR summary preserves as much distributional information as
possible from the original measurement profiles.

A typical use case is CGM data, where each participant contributes many glucose
readings and the goal is to choose thresholds in a data-driven way rather than
relying only on conventional fixed cutoffs (e.g., 54, 70, 181, 251 mg/dL).
The package applies equally to any univariate distributional data from wearable
devices, such as accelerometer activity counts.

**Note:** time-in-range summaries only use measurement values; timestamps are ignored.

## Workflow

The package has two main user-facing functions:

1. `as_distribution()` — preprocess subject-level measurements.
2. `optimal_thresholds()` — estimate data-driven thresholds.

## Step 1: Create a Distribution Object

`as_distribution()` converts raw measurements into an internal representation
(quantile functions on a shared grid) used by the optimizer. It accepts either
a named list of numeric vectors or a long-format data frame.

### List input

```{r}
library(OptiThresholdR)

subjects <- list(
  subject_1 = c(90, 110, 130, 150, 180),
  subject_2 = c(75, 82, 95, 105, 125, 160),
  subject_3 = c(60, 70, 88, 92, 115, 140)
)

distribution <- as_distribution(subjects, range = c(40, 400), M = 200)

distribution
```

### Data frame input

If your data are in data.frame format, pass the subject ID and measurement columns
as strings:

```{r}
df <- data.frame(
  id = rep(c("A", "B", "C"), c(5, 6, 6)),
  gl = c(90, 110, 130, 150, 180,
              75, 82, 95, 105, 125, 160,
              60, 70, 88, 92, 115, 140)
)

distribution_df <- as_distribution(
  df,
  id = "id",
  value = "gl",
  range = c(40, 400),
  M = 200
)
```

### Parameters

- **`range`**: the physical measurement range of the device, e.g., `c(40, 400)`
  for a CGM that records 40–400 mg/dL. If `NULL`, the range is set to the
  observed data range (with a small buffer).
- **`M`**: number of interior grid points to approximate the quantile function. The
  default `M = 200` works well in practice; larger values give finer
  approximations at modest extra cost.

The function `as_distribution` drops missing values automatically and removes subjects with no remaining observations.

## Step 2: Estimate Thresholds

`optimal_thresholds()` searches for `K` thresholds that minimize the chosen
loss function over the measurement range, using differential evolution (DE).

```{r}
fit_loss1 <- optimal_thresholds(
  distribution,
  K = 2,
  loss = "loss1",
  seed = 123
)

fit_loss1
```

### Choosing `K`

`K` is the number of interior thresholds, which partitions the measurement
range into `K + 1` bins. For example, `K = 4` yields 5 TIR proportions, the
same granularity as the standard CGM consensus thresholds (54, 70, 181, 251
mg/dL). Smaller `K` gives coarser but often more interpretable summaries.

### Choosing a loss function

The `loss` argument selects the optimality criterion:

- **`"loss1"`** (distribution preservation): minimizes the average Wasserstein
  distance between each subject's original distribution and its TIR summary.
  Well-suited for characterizing a single, relatively homogeneous cohort.
- **`"loss2"`** (distance preservation): minimizes the discrepancy in pairwise
  subject distances before and after thresholding. Well-suited for
  mixed-population settings where between-group separation matters (e.g.,
  classification or clustering). Requires at least 2 subjects.

```{r}
fit_loss2 <- optimal_thresholds(
  distribution,
  K = 2,
  loss = "loss2",
  seed = 123,
  control = list(NP = 30, itermax = 100, trace = FALSE)   # Custom DEOptim parameters if desired
)

fit_loss2
```


### Interpreting the output

The output object class contains:
- `cutoffs`: the estimated interior thresholds.
- `objective`: the achieved loss value (lower is better).
- `loss` / `wdist`: the loss and distance used for fitting.
- `elapsed`: elapsed optimization time in seconds.

The returned `cutoffs` define `K + 1` bins over the measurement range. Each
subject's TIR summary is the vector of proportions of their measurements
falling within each bin. A lower `objective` means the TIR summary more
faithfully represents the original distributions.

### Solver control

The `control` argument is passed to `DEoptim::DEoptim.control()`. Package
defaults are `NP = max(4, 15 * K)`, `itermax = 1000`, `CR = 0.7`,
`F = 0.75`, `strategy = 3`, `reltol = 1e-4`, `steptol = 30`, `trace = FALSE`.
For quick exploration, reduce `itermax`; for production runs, the defaults
usually suffice. If you supply `control$initialpop`, its row count becomes
the effective `NP`.

## Semi-Supervised Thresholds

In some applications, certain cutoffs are fixed by domain knowledge (e.g.,
70 and 181 mg/dL define the standard CGM time-in-range). The `fixed` argument
keeps those cutoffs in the final set while optimizing `K` additional thresholds
around them.

```{r}
fit <- optimal_thresholds(
  distribution,
  K = 1,        # one additional threshold to optimize
  fixed = 70,   # this cutoff stays in the final threshold set
  loss = "loss1",
  seed = 1,
)

fit
```

Here, `K = 1` means one free threshold is optimized while `70` is kept fixed.
The returned `cutoffs` contain both the fixed and optimized values, sorted.

## Example with Real CGM Data

The [`iglu`](https://irinagain.github.io/iglu/) package provides several
publicly available CGM datasets. Here we use `example_data_hall`, which
contains glucose readings from 5 subjects without diabetes.

```{r, eval = FALSE}
# install iglu if needed (one-time)
devtools::install_github("irinagain/iglu")
```

The dataset `example_data_hall` is in long format with columns including `id` (subject ID)
and `gl` (glucose in mg/dL). The remaining columns are not used by
`as_distribution()`.

```{r, eval = requireNamespace("iglu", quietly = TRUE)}
library(iglu)

data <- example_data_hall
head(data)
```

We set `range = c(40, 400)` to match the Dexcom CGM measurement range and
choose `K = 4` to match the granularity of the standard consensus thresholds
(54, 70, 181, 251 mg/dL). We use `loss2` to capture heterogeneity across subjects.

```{r, eval = requireNamespace("iglu", quietly = TRUE)}
distribution_cgm <- as_distribution(data,
                        id = "id",
                        value = "gl",
                        range = c(40, 400),
                        M = 200)

fit <- optimal_thresholds(distribution_cgm,
                        K = 4,
                        loss = "loss2",
                        seed = 123)

fit
```
